Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-08-10 paper-amia-summit-2018/manuscript.tex
submitted to amia summit 2019

\title{Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions}

@@

Qualitative coding of motivational interviews with pre-defined codes has been traditionally performed manually by trained annotators, which is a tedious and resource-intensive process that involves several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis.\cite{wang2012baselines} In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and analysis \cite{hasan2018predicting} of in-person MI sessions. Specifically, we demonstrated that machine learning methods can be utilized for annotation of MI transcripts according to a simple communication code scheme with the accuracy comparable to human coders.\cite{hasan2016study} Experimental data utilized in these studies, however, were prepared by transcribing audio conversations, which were clearly segmented into utterances by a counselor, a patient and a caregiver.

In this study, we focus on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of motivational interviewing. Specifically, the e-Coaches involved in this study used emails to communicate motivation-enhancing messages that encourage healthy eating among GenY adolescents. E-coaching data is comprised of email responses, which are free-text documents, unlike more traditional dyadic clinical interviews that are naturally segmented into utterances due to their conversational nature.

The unstructured nature of e-Coaching exchanges poses a unique set of challenges for their qualitative analysis. Usually, email segments are automatically coded with the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE), \cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI session. A significant barrier to fully automating the behavior coding process of e-Coaching emails is their segmentation into textual fragments that correspond to distinct communication behaviors. Automating this task is a unique and challenging problem due to the following major reasons:

\begin{enumerate}
\item Emails are unstructured text that contains informal information exchange in a non-traditional format. For example, an e-Coach usually responds to several queries of a patient in one email. However, a provider utterance of MI transcript denotes a particular response for a query of a patient during a traditional MI interview session, which involve a dialogue between a provider and a patient. 
\item Discourse segments in e-Coaching emails do not necessarily correspond to sentences or collection of sentences. One sentence can be segmented into multiple MI behavior fragments. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

Recently, an online clinical intervention called MENU GenY (Making Effective Nutrition Choices for Generation Y) was proposed and evaluated. \cite{alexander2017motivations} MENU GenY is a technology-based public health intervention that relies on personalized e-coaching to encourage increased fruit and vegetable intake among young adults, aged 21-30. The goal of MENU GenY was to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical exchange in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still performed manually, which slows down qualitative analysis of these interventions. This study proposes several machine learning methods to address this problem and the authors are unaware of any other work that focused on the same problem.

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced word embeddings or lexical, punctuation and  part-of-speech (POS) features and experimented with both traditional supervised machine learning method, linear-chain conditional random fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as multilayer perceptrons (MLP),\cite{rumelhart1986learning} bidirectional recurrent neural networks (BRNN)\cite{schuster1997bidirectional} and convolutional recurrent neural networks (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers of electronic health records (EHR)\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection of EHR.\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence} Apostolova et al. applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports in EHR into sections, such as demographics, history, procedure, finding and impression.\cite{apostolova2009automatic} After identification of each line in the document, Tepper et al. trained maximum entropy models to classify the document into pre-defined sections, such as general patient information, provider information, medical condition, medical history, clinical history, findings, impression, etc.\cite{tepper2012statistical} Denny et al. proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and a Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. \cite{denny2009evaluation} On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries in EHR. Liu et al. have shown that a linear chain CRF performed better than the hidden markov and maximum entropy model in sentence boundary detection, which was evaluated using the official NIST evaluation tools.\cite{liu2005using} Recently, Treviso et al. proposed a recurrent convolutional neural network for automatic identification of sentence boundary in a neuropsychological text in the Portuguese language.\cite{treviso2017sentence} Their proposed model combined prosodic and part of speech features with pre-trained GloVe word embeddings,\cite{pennington2014glove} which achieved 0.74 F1-score in sentence boundary detection on the Brazilian constitution dataset. Segmentation of e-Coaching emails is different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers. Segmentation of e-Coaching emails corresponds to a special type of discourse analysis \cite{webber2012discourse} aimed at better understanding the effective e-Coaching communication strategies and revealing the unique socio-psychological characteristics of a patient.

Our proposed approach is novel for the segmentation of e-Coaching emails since previous studies mainly focused on the segmentation of EHR documents into sections, headers and sentences. However, this study segmented clinical exchange into groups of MI behaviors which will significantly reduce the amount of resource and time required to segment clinical exchange manually. Furthermore, these segmentation models could be integrated with auto coding classifiers to build a software pipeline for automated annotation of clinical exchanges.

\section*{Methods}
\subsection*{\textit{Data collection}}

The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We formulate the segmentation task as a binary classification problem where each word or punctuation mark annotated with a label 1 or 0, to indicate whether it precedes a new segment or not. In total, we obtained 95,777 words and 7,140 punctuation marks, which include 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning method, conditional random fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as multilayer perceptrons (MLP),\cite{rumelhart1986learning} bidirectional recurrent neural networks (bidirectional RNN or simply BRNN)\cite{schuster1997bidirectional} and convolutional recurrent neural networks (CRNN).\cite{treviso2017sentence} For MLP models, instances were created with $2l$ words or punctuations at each step in a word sequence, where an instance contains next $l$ words or punctuations and prior $l$ words or punctuations including current word or punctuation. Each instance is classified into either ``new segment'' or ``same segment'' based on whether the $l^{th}$ word or punctuation in the gold standard precedes a new segment or not. For CRF, BRNN and CRNN models, an email was taken as input sequence, such that POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of the model at each step. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora.\cite{pennington2014glove, mikolov2013distributed} This study utilized embeddings estimated on Google News corpus consisting of 1.6 billion words referred to pre-trained word2vec word embeddings.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} When words or punctuation marks are not found in the pre-trained word vectors, word embeddings trained with e-Coaching email corpus referred to trained word embeddings were utilized. CRF utilized lexical features which represent each word in an e-Coaching email.

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three deep learning models. Since deep learning methods provide a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\textbf{Conditional Random Fields}: CRF has been widely used in various NLP tasks including part-of-speech tagging and segmentation tasks.\cite{lafferty2001conditional, hirohata2008identifying} Unlike a maximum entropy markov model which uses per-state exponential models for the conditional probabilities of next states given the current state, a CRF is modeled with an exponential distribution provides the conditional distribution of the entire output sequence given the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ for output and input sequences, $y$ and $x$, by the Eq. \ref{Eq:crf}

@@

Here $Z_x$ is a normalization factor over all possible labelings of $x$, and $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently by the Viterbi algorithm. In our experiments, the following features were utilized with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating the word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc) or not iv) binary feature indicating the word is a title word or not? (e.g. ``The'' is a title word but ``the'' is not) v) POS tags

\textbf{Multilayer Perceptrons}: MLP is a feed-forward artificial neural network which maps set of input onto a set of appropriate output.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} shows a multilayer perceptron with a single hidden layer. In MLP networks, there is no cycle or loop and information moves forward only, from the input nodes through the hidden nodes and to the output nodes. This study utilized MLP with a nonlinear activation function ($relu$) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we implement dropout on the last layer, which randomly hides some neurons during the training phase. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models.

@@

\textbf{Bidirectional Recurrent Neural Networks}: BRNN is a neural network designed to capture sequential patterns by considering both past \& future inputs as well as complex relationships between input features and output labels.\cite{schuster1997bidirectional} The output of BRNN layer is computed with the summation of the forward RNN output and backward RNN output. Gated Recurrent Units (GRU)\cite{chung2014empirical} are utilized as RNNs that are capable of handling variable size input sequence and have an internal memory that can be reset. Figure~\ref{fig:crnn} will represent an architecture of BRNN if convolution layer is removed.

\textbf{Convolutional Recurrent Neural Networks}: CRNN is a deep neural networks architecture,\cite{treviso2017sentence} shown in Figure~\ref{fig:crnn}, consisting of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input layer. The embedding layer provides a distributed representation of words, gives us $m \times n_e$ matrix after fetching the pre-trained word vectors. When POS tags are used in combination with word embeddings, 10-dimensional POS vectors are concatenated with 300-dimensional word vectors to obtain new vectors $n_e = n_w + n_p$ of size 310. The convolution layer used this word embeddings to extract new features depending on neighbor words. This study used 1D convolution which simply moves one step at a time on vertically. In this layer, one filter is responsible for the extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the text, $n_f$ useful features are produced in the convolution layer for each word. A max pooling operation was then performed over time to find the most significant features in a text region. The bidirectional recurrent layer receives new features extracted in the convolution layer. RNN is usually used to capture the previous history. Moreover, bidirectional RNN is capable of capturing both previous and past histories because bidirectional RNN utilized the features by looking at forward and backward states. The fully connected layer used the bidirectional RNN output for classifying each word or punctuation into ``new segment'' and ``same segment'' classes. Since a fully connected layer has a greater number of parameters, they are more likely excessively co-adapt these parameters causing over-fitting. To prevent this over-fitting problem, dropout is used to randomly drop 50\% of the connections on the fully connected layer. After that, a sigmoid operation is performed on the output layer to provide the probability of whether the word or punctuation precedes a new segment or not. We experimentally determined the optimal parameters and found that the best performance is achieved with 5 folds cross-validation when filter length in convolution layer is 7, number of filters is 100, max-pool size is 3, convolutional layer activation is $relu$, RNN layer activation is $tanh$ and number of recurrent units in RNN is 200. Adam\cite{kingma2014adam} is used for optimization and the early stopping strategy is applied with 50 epochs when the batch size is 32 and the learning rate is 0.001. To utilize our source codes by other research studies, source codes of all experiments will be publicly available\footnote{https://github.com/teanalab/xxxxx} upon acceptance of this paper for publication.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Architecture of convolutional recurrent neural networks for automatic segmentation of e-Coaching emails}}
    \label{fig:crnn}
\end{figure}

\subsection*{\textit{Evaluation metrics}}
We report standard metrics for experiments (precision, recall and F1-measure) to evaluate the performance of binary classifiers.\cite{aas1999text} However, accuracy is not reported as a performance metric because accuracy is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem for an unbalanced dataset. The results are reported based on 5 folds cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets. \cite{hu2015kernelized}

\section*{Results}
Our experimental results have four important directions. First, the optimal word embedding dimension and instance size of MLP models are illustrated in Figure~\ref{fig:embedding-dimension-mlp}. Second, results are reported with respect to ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes in Table~\ref{tab:result_base}. Third, classification performance of different machine learning methods are outlined in Table~\ref{tab:result_weighted_avg} when word embeddings or lexical features are used in combination with punctuation and POS features. Fourth, the impact of individual features, as well as their combination on the task of email segmentation, are demonstrated in Table~\ref{tab:result_roc}.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/mlp-and-vector.eps}
    \caption{\textbf{Performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe and word2vec models (left). Performance of MLP model on e-Coaching email segmentation by varying the length of instance (right).}}
    \label{fig:embedding-dimension-mlp}
\end{figure}

Figure~\ref{fig:embedding-dimension-mlp} (left) shows the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe\footnote{https://nlp.stanford.edu/projects/glove/} and word2vec models. It was observed that the best performance of email segmentation was achieved with pre-trained 300-dimensional word2vec word vectors when three types of features are used together. Therefore, results of all deep learning models are reported with pre-trained word2vec 300-dimensional word vectors. In MLP, inputs were prepared by computing the summation of first $l$ word vectors and last $l$ word vectors where each instance contains $2l$ words or punctuations. Figure~\ref{fig:embedding-dimension-mlp} (right) demonstrates the performance of MLP models on e-Coaching email segmentation by varying the value of $l$. It was observed that the best performance of MLP classifiers is achieved when $l = 2$. Therefore, results of all MLP models are reported with $l = 2$.

As follows from Table~\ref{tab:result_base}, CRNN outperforms all other models in terms of recall and F1-measure achieving 0.797 recall with 0.785 F1-measure for new segment detection. CRNN also shows superior performance in all performance metrics for weighted average over ``new segment'' and ``same segment'' classes. BRNN demonstrates the lowest performance among all models in terms of precision and F1-Measure. On the other hand, MLP has the highest precision of 0.836 when word embeddings or lexical features are used to identify ``new segment''. CRF achieves 0.733 F1-Measure in new segment class and 0.984 F1-Measure in overall, which demonstrates the second highest performance to identify ``new segment'' as well as a weighted average over both classes. Experimental results indicate that performance of all classifiers as a weighted average over both classes is remarkably higher compared to ``new segment'' class, which is expected since 96.95\% instances belong to the ``same segment'' categories and 99.3\% of them correctly classified. For example, as a weighted average over ``new segment'' and ``same segment'' classes, CRNN achieves 27.23\%, 23.71\% and 25.61\% higher precision, recall and F1-measure, respectively, compared to ``new segment'' detection. \\

>>>>

\title{Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions}

@@

Qualitative analysis of motivational interviews has been traditionally performed manually by MI researchers, which is a tedious and resource-intensive process involving several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis.\cite{wang2012baselines} In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and predicting the outcome \cite{hasan2018predicting} of in-person MI sessions. Experimental data utilized in these studies, however, were transcribed auto recordings of in-person MI sessions with a counselor, which have a clear discourse structure and were clearly segmented into the utterances by a counselor, a patient or a caregiver.

In this study, we focus on the analysis of e-Coaching sessions, behavioral interventions that are delivered via email and are grounded in the principles of MI. Specifically, the e-Coaches involved in this study used emails to communicate motivation-enhancing messages that encourage healthy eating among GenY adolescents. E-coaching data is comprised of email responses, which are free-text documents, unlike more traditional dyadic clinical interviews that are naturally segmented into utterances due to their conversational nature.

The unstructured nature of e-Coaching exchanges poses a unique set of challenges for their qualitative analysis. Usually, email segments are automatically coded with the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE), \cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI session. A significant barrier to qualitative analysis of e-Coaching exchanges is their segmentation into textual fragments that correspond to a distinct e-Coach and patient communication behaviors. Automating this task is a unique and challenging problem due to the following reasons:

\begin{enumerate}
\item Emails are unstructured text containing informal information exchange in a non-traditional format. For example, an e-Coach usually responds to several previous patient statements in one email. However, in a traditional MI interview session involving a dialogue between a provider and a patient, a provider utterance is a response to an immediately preceding inquiry or statement by a patient. 
\item Discourse segments in e-Coaching emails do not necessarily correspond to sentences or paragraphs. One sentence can be divided into fragments corresponding to multiple MI behaviors. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/segment-example.png}
    \caption{\textbf{Example of e-Coaching emails segmented into fragments that correspond to MI behaviors of an e-Coach and a patient}}
    \label{fig:text-segment}
\end{figure}

Figure~\ref{fig:text-segment} illustrates a segmentation of an e-Coaching email exchange, in which the first sentence is segmented into 2 MI behavior fragments, while the fourth and fifth segments correspond to one and three sentences, respectively. Segmentation of e-Coaching email exchanges between patients and e-Coaches corresponds to a special case of discourse analysis \cite{webber2012discourse} aimed at better understanding the effective communication strategies specific to this type of behavioral interventions and revealing the unique socio-psychological characteristics of patients.

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of qualitative analysis of this type of clinical communications. Specifically, for this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a recent technology-based public health intervention that relies on personalized e-Coaching to encourage increased fruit and vegetable intake among young adults, aged 21--30. The goal of MENU GenY was to develop a better coding scheme for e-Coaching communications aimed at improving GenY eating habits. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail based behavioral interventions}.

Specifically, we introduce and evaluate the effectiveness of word embedding or lexical, punctuation and part-of-speech (POS) features in conjunction with both traditional supervised machine learning methods, such as linear-chain Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning architectures, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

\section*{Relevant work}
Previous relevant work in the biomedical domain primarily focused on segmentation of text in electronic health records (EHR) into sections and headers \cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection.\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence} In particular, Maximum Entropy models \cite{tepper2012statistical} and Support Vector Machine (SVM) along with word-vector cosine similarity metric and several heuristics \cite{apostolova2009automatic} have been applied to segment or classify clinical documents in EHR into pre-defined sections, such as general patient information, medical history, procedures, findings, etc. Denny et al. \cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and a Naive Bayes classifier to identify sections and headers in EHR. Segmentation of e-Coaching emails, however, is different from segmentation of other clinical documents in that the focus is on clinical conversation.

SVM in conjunction with prosodic and part of speech features \cite{kreuzthaler2015detection} and Recurrent Convolutional Neural Networks \cite{griffis2016quantitative} have also been utilized for \textit{sentence boundary detection} in general text. Liu et al. \cite{liu2005using} demonstrated that a linear-chain CRF outperforms Hidden Markov and Maximum Entropy models for this task.

Segmentation of e-Coaching emails is different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers. The proposed methods will automate the process of segmenting clinical exchanges into groups of MI behaviors, which will significantly reduce the resources and time required to perform this task manually. Furthermore, these methods can be integrated with auto coding methods \cite{hasan2016study,kotov2015interpretable} to create a software pipeline for automated analysis of clinical conversation.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We consider segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a label of 1 or 0 to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing instances were created based on a sliding window of $2n$ words or punctuation marks over each position in a given input sequence, such that each instance consists of a pair of the next $n$ words or punctuation marks and prior $n$ words or punctuation marks, including the current word or punctuation mark. Each sample is classified into either a ``new segment'' or a ``same segment'' based on whether there should be a segment break after the current word or not. In the case of CRF, BRNN and CRNN models, an email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output of a model. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic units of semantics. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embedding is one such representation, when each word is associated with a real-valued vector in a high dimensional vector space. Distributed representations of words have been shown to capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized embeddings estimated on Google News corpus consisting of 1.6 billion words pre-trained using word2vec word embedding method.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} When words or punctuation marks are not found in the pre-trained word vectors, we utilized word embeddings trained with our e-Coaching email corpus, which we refer to as corpus-based embeddings. CRF utilized lexical features, which correspond to words in e-Coaching emails.

@@

We experimented with 4 different classifiers, including one traditional machine learning method, CRF and three deep learning models. Since deep learning architectures provide a flexible mechanism for constructing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\textbf{Conditional Random Field}: CRF has been widely used in various NLP tasks, such as part-of-speech tagging.\cite{lafferty2001conditional, hirohata2008identifying} Unlike a maximum entropy Markov model, which uses per-state exponential models for conditional probability of the next state given a current state, CRF model directly estimates conditional distribution of the entire output sequence, given the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ for output and input sequences, $y$ and $x$:

@@

where $Z_x$ is a normalization factor, $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently using the Viterbi algorithm. In our experiments, the following features were utilized in conjunction with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating whether a word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc.) or not iv) binary feature indicating whether a word is a title word or not (e.g. ``The'' is a title word but ``the'' is not) v) POS tags

@@

\textbf{Multi-Layer Perceptron}: MLP is a feed-forward artificial neural network, which maps an input onto one or several outputs.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} shows a multilayer perceptron with a single hidden layer. In MLP networks, there is no cycle or loop and information moves forward only, from the input nodes through the hidden nodes and to the output nodes. This study utilized MLP with a nonlinear activation function (rectified linear unit) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we utilized dropout in fully connected layers by randomly hiding neurons during training. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models.

\textbf{Bidirectional Recurrent Neural Network}: BRNN is a neural network designed to capture sequential patterns by considering both past \& future inputs as well as complex relationships between input features and output labels.\cite{schuster1997bidirectional} The output of BRNN layer is computed as an aggregation of outputs of the forward and backward RNNs. Gated Recurrent Units (GRU)\cite{chung2014empirical} capable of handling variable size input sequence and having internal memory, which can be reset, were utilized as an RNN. Figure~\ref{fig:crnn} represents the architecture of BRNN, in the case when a convolution layer is removed.

\textbf{Convolutional Recurrent Neural Network}: CRNN is a deep neural network architecture,\cite{treviso2017sentence} shown in Figure~\ref{fig:crnn}, which consists of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. 
E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input layer to produce a $m \times n_e$ matrix after fetching the pre-trained word vectors. When POS tags are used in combination with word embeddings, 10-dimensional POS vectors are concatenated with 300-dimensional word vectors to obtain new vectors $n_e = [n_w;n_p]$ of size 310. The primary purpose of convolution is to extract new features depending on the neighboring words. We used 1D convolution. In this layer, one filter is responsible for extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the input text, $n_f$ useful features are produced in the convolution layer for each word. A max pooling operation was then performed over time to find the most significant features in a textual fragment. The bidirectional recurrent layer receives new features extracted in the convolution layer. RNN is usually used to capture long-range dependencies in a sequence of observations. Moreover, bidirectional RNNs are capable of capturing both past and future contexts through forward and backward traversals of a sequence. The purpose of the fully connected layer is to use the output of bidirectional RNN layer for classifying each word or punctuation into ``new segment'' and ``same segment'' classes. Since a fully connected layer has a greater number of parameters, they are more likely to excessively co-adapt with these parameters, which may cause over-fitting. To prevent this, we utilized dropout by randomly disregarding 50\% of the connections to the fully connected layer. After that, logistic sigmoid is applied to the output layer to provide the probability of whether the word or punctuation precedes a new segment or not. We experimentally determined the optimal parameters and found that the best performance is achieved with 5-fold cross-validation when filter length in convolutional layer is 7, number of filters is 100, max-pool size is 3, ReLU is used as a convolutional layer activation function, RNN layer activation is hyperbolic tangent and the number of recurrent units in RNN is 200. Adam\cite{kingma2014adam} with 50 epochs, the batch size of 32 and learning rate of 0.001 was used for optimization and the early stopping strategy was applied.\footnote{source code will be available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Architecture of convolutional recurrent neural networks for automatic segmentation of e-Coaching emails}}
    \label{fig:crnn}
\end{figure}

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric because it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5 fold cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets. \cite{hu2015kernelized}

\section*{Results}
Our experiments span four dimensions. First, we report the optimal size of word embeddings and sliding window size of the MLP model. Second, results are reported with respect to ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes in Table~\ref{tab:result_base}. Third, classification performance of different machine learning methods are summarized in Table~\ref{tab:result_weighted_avg} when word embeddings or lexical features are used in combination with punctuation and POS features. Fourth, the impact of the individual features, as well as their combination on the segmentation task, is reported.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/mlp-and-vector.eps}
    \caption{\textbf{Performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and corpus word embeddings with GloVe and word2vec models (left). MLP model on e-Coaching email segmentation by varying the sliding window size $2n$ (right).}}
    \label{fig:embedding-dimension-mlp}
\end{figure}

Figure~\ref{fig:embedding-dimension-mlp} (left) shows the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and corpus word embeddings with GloVe \footnote{https://nlp.stanford.edu/projects/glove/} and word2vec models. It was observed that best performance is achieved with pre-trained 300-dimensional word2vec word vectors when three types of features are used together. Therefore, we report our results with word2vec 300-dimensional word vectors for all deep learning models used in this study. For MLP models, the first layer input was prepared by the summation of the first $n$ word vectors in the sliding window and the last $n$ word vectors when a sliding window contains $2n$ words or punctuations. Figure~\ref{fig:embedding-dimension-mlp} (right) demonstrates the performance of the MLP model on e-Coaching email segmentation by varying the size of the sliding window. It can be observed that the best performance of MLP is achieved when $n$ is 2. Therefore, results of MLP in the remaining experiments are reported with $n$ set to 2.

As follows from Table~\ref{tab:result_base}, CRNN outperforms all other models in terms of recall and F1-measure achieving 0.797 recall with 0.785 F1-measure for new segment detection. CRNN also shows superior performance in all performance metrics for weighted average over ``new segment'' and ``same segment'' classes. BRNN demonstrates the lowest performance among all models in terms of precision and F1-Measure. On the other hand, MLP has the highest precision of 0.836 when word embeddings or lexical features are used to identify ``new segment''. CRF achieves 0.733 F1-Measure in new segment class and 0.984 F1-Measure overall, which corresponds to the second highest performance in identifying ``new segment'' as well as a weighted average over both classes. Experimental results indicate that performance of all classifiers as a weighted average over both classes is remarkably higher compared to ``new segment'' class, which is expected since 96.95\% of instances belong to the ``same segment'' class and 99.3\% of them are correctly classified. For example, as a weighted average over ``new segment'' and ``same segment'' classes, CRNN achieves 27.23\%, 23.71\% and 25.61\% higher precision, recall and F1-measure, respectively, compared to the ``new segment'' detection.
